\section{What is this, why does this exist}
It would be a good idea to read the previous chapter on Simplex to refresh memory of what it does. 

But OCR MEI is needy, so we also need to know how we can use LP to model and solve real-world problems (so they say...) that involve optimisation. It's actually pretty cool, but can also get a bit boring. 

You need to be able to apply to LP to these four areas, so it's ultimately good revision! 

\begin{enumerate}
    \item Shortest path 
    \item Longest path 
    \item Capacity networks
    \item Matching problems 
    \item Allocation 
    \item Transportation 
\end{enumerate} 

The best way to practice this to topic is.. well - to just do shed loads of questions on them. MEI can be pretty harsh on them, so doing questions is the way to go as boring as it sounds!

\section{Indicator Variables}


\section{Network Flows}
This is a bit different from shortest/longest paths. In this case, the variables are the flows through the arcs. 

So how the fuck do you deal with this if you're not using indicator variables? 
\begin{itemize}
    \item Maximise flows going out of S 
    \item Maximise flows going into T
\end{itemize}

Consider this flow below: 

Out objective function here could be 

$$
SA + SB
$$

Subject to: this is flow going it take away the flow out = 0 
$$
SA + BA - AB - AT - AT = 0 
$$
$$
SB + AB - BA - BT = 0 
$$
$$
SA \leq 8 
$$
$$
SB \leq 5
$$
$$
AB \leq 7 
$$
$$
BA\leq 7 
$$
...etc 


\subsection{Finding maximum flows}
Either look at sum of edges come out of S, or sum of edges going into T


\section{CPA}

WE want to minimise the time taken to complete a task. 


\section{Summary}
I've had to just do a summary! (It's getting late and I ought not be working too late into my bedtime and (2) - best way like I said to actually do questions on these topics) 

\begin{table}[h!]
\centering
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{|p{3.5cm}|p{3.5cm}|p{3.5cm}|p{4cm}|}
\hline
\textbf{Problem} & \textbf{Objective} & \textbf{Variables} & \textbf{Key Constraints} \\
\hline
\textbf{Shortest Path} & Minimise total distance or cost & $x_{ij} \in \{0,1\}$ (or $\geq 0$ in LP relaxation) & Flow conservation at all nodes; 1 unit out from start, 1 unit into end \\
\hline
\textbf{Longest Path (CPA)} & Maximise event time at the end node & $d_i \in \mathbb{R}$ (event start times) & $d_j \geq d_i + t_{ij}$ for each arc $i \to j$ \\
\hline
\end{tabular}
\caption{LP Formulations for Shortest and Longest Path Problems}
\end{table}



\begin{table}[h!]
\centering
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{|p{3cm}|p{3.5cm}|p{3.5cm}|p{4cm}|}
\hline
\textbf{Application} & \textbf{What’s Being Optimised?} & \textbf{LP Objective} & \textbf{Key Constraints} \\
\hline
\textbf{Network Flow} & Total flow from source to sink & Maximise flow & Arc capacities, flow conservation \\
\hline
\textbf{Allocation} & Task performance, resource use & Maximise benefit or minimise cost & Resource availability \\
\hline
\textbf{Transportation} & Goods delivery & Minimise cost & Supply and demand \\
\hline
\textbf{Matching} & Pairings between elements & Maximise weight / minimise cost & Each item matched once \\
\hline
\end{tabular}
\caption{Summary of Linear Programming Applications}
\end{table}

